---
title: "finalProject"
output:
  pdf_document: default
  html_document: default
date: "2023-04-09"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(rvest)
library(dplyr)
library(readxl)
library(tree)
library(rpart)
```

## 
1. Reading in the data 
2. Cleaning the data
3. Merging the data
4. Initial analysis/ Graphing
  - Correlation of all to salary
  - Correlation between predictors
5. Initial modeling techniques
6. Cross validation and model evalulation techniques
7. Separating Salary Outliers and retesting the models to check performance
8. TBD


## 1. Reading in the Data
```{r Reading in the Data}
url_salary <- 'https://hoopshype.com/salaries/players/2021-2022/'
x_path <- '//*[@id="content-container"]/div/div[3]/div[2]/table'
salary_table <- url_salary %>%
  read_html() %>%
  html_element(xpath = x_path) %>%
  html_table()

df.stats <- read_excel('Player_stat_data.xlsx')
```

## 2. Cleaning the data
```{r Cleaning the dataframes}
# For Salary
colnames(salary_table) <- as.character(salary_table[1,])
salary_table <- salary_table[-1,]
salary_table <- subset(salary_table, select  = -ncol(salary_table))
salary_table <- subset(salary_table, select = -which(names(salary_table) == "NA"))
salary_table$`2021/22`<- gsub("\\$", "", salary_table$`2021/22`)
salary_table$`2021/22`<- gsub(",", "", salary_table$`2021/22`)
names(salary_table)[names(salary_table) == "2021/22"] <- "salary"
salary_table$salary <- as.numeric(salary_table$salary)
                                
# For Player Stats
age <- select(df.stats, Age) 
df.stats <- select(df.stats, -Age)
tm <- select(df.stats, Tm) 
df.stats <- select(df.stats, -Tm)
df.stats.distinct <- df.stats %>% 
  distinct(Player, .keep_all = TRUE)
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "FG%"] <- "FGPerc"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "3P%"] <- "ThreePPerc"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "2P%"] <- "TwoPPerc"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "eFG%"] <- "eFGPerc"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "FT%"] <- "FTPerc"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "3P"] <- "ThreePointers"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "2P"] <- "TwoPointers"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "3PA"] <- "ThreePointAvg"
colnames(df.stats.distinct)[colnames(df.stats.distinct) == "2PA"] <- "TwoPointAvg"
```

## 3. Merging the data
```{r Merging the Data}
merged.player.data <- merge(salary_table, df.stats.distinct, by = "Player", all = FALSE)

```



## 4 Exploratory Data Analysis
Plan:
- Look to graph all the features to understand their distribution and relation to salary
- Use histogram and plot to do this, and look up for some more creative plots: https://r-graph-gallery.com/ggplot2-package.html
- Build a correlation matrix of the data with each other and with salary using heatmap: https://r-graph-gallery.com/heatmap

https://cran.r-project.org/web/packages/vtable/vignettes/sumtable.html for summary stats
- 
```{r EDA}

plot(merged.player.data$FG, merged.player.data$salary)
abline(lm(merged.player.data$salary ~ merged.player.data$FG))
plot(merged.player.data$eFGPerc, merged.player.data$salary)
hist(merged.player.data$eFGPerc)
hist(merged.player.data$TwoPPerc)
hist(merged.player.data$ThreePPerc)
hist(merged.player.data$salary)
```

```{r}
merged.player.data$log_salary <- log(merged.player.data$salary)
hist(merged.player.data$log_salary)
```

## 5 Modeling
- Tree
- Logistic Regression
- KNN
- more


```{r Training Test Data Creation}
player.col <- merged.player.data[,1]
merged.player.data <- merged.player.data[,-1]
sal.col <- merged.player.data[,1]
merged.player.data <- merged.player.data[,-1]
data.size<-nrow(merged.player.data)
train.rows<-sample(1:data.size, data.size/2)
train.data<-merged.player.data[train.rows,]
test.data<-merged.player.data[-train.rows,]
true.vals<-test.data[,28]
```

```{r}
tree.salary <- tree(log_salary ~., train.data)
summary(tree.salary)

plot(tree.salary)
text(tree.salary)
```

```{r}
tree.pred <- predict(tree.salary, test.data)
mse <- mean((tree.pred - true.vals)^2)
mse
```

```{r}
cv.salary<-cv.tree(tree.salary)
plot(cv.salary$size, cv.salary$dev, type="b")
```

```{r}
best.size <- 3
pruned <- prune.tree(tree.salary, best = best.size)
plot(pruned)
text(pruned)
```


```{r}
prune.pred <- predict(pruned, test.data)
mse <- mean((prune.pred - true.vals)^2)
mse
```

The pruned model does not perform better than the unpruned model, as the mse was 1.112142 while the unpruned model had a mse of 1.068486. Therefore, we will stick with the unpruned model as the best version of the tree model. 

## 6 Removing salary outliers and rerunning models 

## 7 Comparing Model performance

jon added
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
